---
title: "Homework 4"
subtitle: "Discrete Random Variables"
author: "WRITE YOUR NAME HERE"
output: 
  pdf_document: 
        highlight: null
        pandoc_args: ["--syntax-highlighting=tango"]
header-includes:
    - \usepackage{amsmath, amssymb}
    - \usepackage{setspace} 
---
\onehalfspacing

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
#R packages
 

```

* * * 

### Instructions

- This homework is due in Gradescope on Wednesday May 6 by midnight PST. There is a 15 minute grace period and submissions made during this time will not be marked as late. Any work submitted past this period is considered late.  

- Please answer the following questions in the order in which they are posed. \textbf{Be sure to support your work and show your steps. All calculations must be done within your document and shown with the answer printed}

- Don't forget to (i) make a local copy this document for your work and to (ii) knit the document frequently to make sure there are no compilation errors. 

- When you are done, download the PDF file as instructed in section and submit it in Gradescope. \textbf{Do not forget to tag your answers to every sub-part of every question. We will take off 10\% for every part that is not tagged}

- \textbf{Please read the acceptable policies for this class in CANVAS. In particular, you are NOT ALLOWED to consult external sources - AI, tutors, tutoring sites, internet - to complete any part of this work. Do not use notation or definitions or results we have not covered in this class. We will deduct points for this and/or report the matter for further investigation. }

* * *


### Exercises 


1. (School bus) Four buses carrying 148 students from the same
school arrive at a football stadium. The buses
carry, respectively, 40, 33, 25, and 50 students. 

<!--- Note for 1a, 1b: please consult chapter 5 slidedeck for more about PMFs. In particular, you only need to show the values which have positive probability. The first column should list the probability (leave it as a fraction) and the second column should show the possible values --->

a. Suppose one of the students is randomly selected. Let $X$ denote
the number of students that were on the bus carrying the randomly selected student. Replace the blanks to show the possible values and their respective probabilities.  Then write a couple of sentences explaining your calculation.

$$f_1(x) = \begin{cases}
          ---,  &  x = ---  \\[3pt]
          ---,  & x = --- \\[3pt]
          ---,  & x = --- \\[3pt]
          ---,  & x = ---
         \end{cases}$$
         
b. Suppose one of the 4 bus drivers is also randomly selected. Let the random variable $Y$ denote the number of students on their bus. Replace the blanks to show the possible values and their respective probabilities.  Then write a couple of sentences explaining your calculation.

$$f_2(y) = \begin{cases} 
          ---,  & y = ---  \\[3pt]
          ---,  & y = --- \\[3pt]
          ---,  & y = --- \\[3pt]
          ---,  & y = ---
         \end{cases}$$

c. Which do you think will be larger typically? Answer $X$ or $Y$ and give a brief explanation of your thinking.

2. (Bases) The CDF of the number of bases reached by a baseball player is shown below.

$$F(x) = \begin{cases}
            0 & x < 0 \\[2pt]
            0.513 & 0 \leq x < 1 \\[2pt] 
            0.856 & 1 \leq x < 2 \\[2pt]
            0.899 & 2 \leq x < 3 \\[2pt]
            0.911 & 3 \leq x < 4 \\[2pt]
            1     & 4 \leq x 
          \end{cases}$$

<!--- Note for 2a: see Chapter 5 slidedeck on slide 33 for the properties. You have to make sure each one is satisfied. Also see problem 3 from section in week 4 for something similar  --->   

a. Argue that $F(x)$ is a valid CDF.

<!--- Note for 2b: use the cases environment as shown in problem 1. If you find yourself reaching for a calculator, stop! Make a code chunk and do your calculations in it  --->

b. Write $f(x)$, the PMF of $X$ below. No explanation needed.


<!--- Note for 2c: show the probability as expressions that involve $f$ and $F$ first. Then calculate both in a code chunk with code shown and answers outputted so we/you can verify they are the same. See example 5.6 from chapter 5 slidedeck
--->

c. Find $P(1.5 \leq X < 4)$ in two ways: one using the PMF and the other using the CDF. 

3. (Capture-recapture) From a lake containing 200 goldfish, a sample of 25 fish are caught, tagged and released. Later, a second sample of 20 fish are caught randomly and the number of tagged fish in the sample are counted.

<!--- Note for 3a: this is again like the acceptance sampling scenario. Think about why. Also see example 5.4 from the Chapter 5 slidedeck for what it means to write the PMF as a formula. 

This method is called capture-recapture and it is used to estimate wildlife populations. We assume that the population is closed: in other words there are no additions (breeding, immigration) or losses (deaths) in between the capture and recapture. 
--->

a. Let $X$ denote the number of tagged fish in the second sample. Write the PMF of $X$ as a formula. No explanation is needed.   


<!--- Note for 3b: look at example code on slide 25 of chapter 5. Don't forget the `labs` layer for ggplot, and please give a descriptive title that sets the context for what the graph is showing, not just "Problem 3b". Also be sure to use the `scale_x_continuous` layer so the ticks are placed at the possible x values. See section from week 4 if you need help. --->

b. Create a probability histogram of $X$. 

    ```{r prob-hist, fig.alt = " give alternative text for your image here to make it accessible to screen reader", out.width = "50%", fig.align = "center"}
    
    #attach tidyverse in setup 
    
    ```

c. Comment on the shape of the distribution from part b, the location of the peaks and why it makes sense.


<!--- For 4: In Roulette, bets on color (red or black) are 1 for 1 or  even money. That is, say you pay \$1 to play. If your color shows up, you get \$2. So you win \$1. If you lose, they keep your money. So you win -\$1. --->


4. (Roulette) A gambling book recommends this "winning" strategy for the game of roulette: 

    - Bet \$1 on red. If red appears (which has probability $\frac{18}{38}$ ), then take the \$1 profit and quit. 
    
   - If red does not appear (which has probability $\frac{20}{38}$ of occurring), make additional \$1 bets on red on each of the next two spins of the roulette wheel and then quit.
    
    Let the random variable $X$ denote your winnings when you quit. Also denote $E_i$ as the event that red appears on the ith spin, where $i$ can be 1, 2 or 3. 

a. What are the possible values for $X$? For each possible value, describe it in terms of the $E_i$. Give a brief explanation.

<!--- Hint for 4b:  Write your answer in terms of probabilities of events first, then do the calculation in a code chunk.  --->


b. Find $P(X > 0)$. You may assume the outcomes from the spins are mutually independent, which incidentally also implies that their complements are independent.

