---
title: "Problem Section 6"
subtitle:  "Expected Value and Variance"
graphics: yes
output: 
      pdf_document:
      highlight: null
      pandoc_args: ["--syntax-highlighting=tango"]
header-includes:
    - \usepackage{amsmath, amssymb}
    - \usepackage{framed}\definecolor{shadecolor}{rgb}{0.949,0.949,0.949}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

* * * 
\begin{shaded}

\textbf{Learning Outcomes}

The problems are designed to build conceptual understanding and problem-solving skills. The emphasis is on learning to find, evaluate and build confidence.
The specific tasks include: 

   - Calculate expected value and variance using the PMF
     
   - Use Chebychev's inequality to make probability statements
 
   - Back up and support work with relevant explanations

 

\end{shaded}


* * *

### Exercises 

1. (Law of the Unconscious Probabilist) Weird Wally offers you the following choice. You may have a flat amount of $\frac{1}{3.5}$ dollars or you may roll a fair die and he will give you $\frac{1}{X}$ dollars where $X$ is the value on the roll. 
The PMF of $X$ is shown in the table below.

    \begin{table}[h]
    \centering
    \begin{tabular}{c|c|c|c|c|c|c} \hline
    $x$ & 1& 2 & 3 &4 & 5& 6 \\ \hline
    $f(x)$& $\frac{1}{6}$ & $\frac{1}{6}$ & $\frac{1}{6}$ &$\frac{1}{6}$ &$\frac{1}{6}$     &$\frac{1}{6}$ \\ \hline
    \end{tabular}
    \end{table}

    Which is the better deal? 


    In order to judge, we will calculate $E\left[ \frac{1}{X} \right]$. We can find it by using the Law of the Unconscious Probabilist:

    \begin{align*}
    E\left [\frac{1}{X}\right ] &= \sum_x \frac{1}{x} \cdot f(x) \\ 
    &= \frac{1}{6}\sum_{x=1}^6\frac{1}{x}
    \end{align*}

    We may calculate this using R.

    ```{r}
    inv_x <- 1/(1:6)
    E_1_x <- sum(inv_x)/6
    E_1_x
    ```

    Using R we see that $E\left[ \frac{1}{X} \right ] = `r E_1_x`$ versus Wally's offer of $\frac{1}{3.5} = `r 1/3.5`$. Thus rolling the die has a better expected value, and this is the better deal. 
    
2.  A shipment consists of hundred memory chips, 4% of which are defective. A random sample of twenty chips is
selected from the shipment.
Let $X$ denote the number of defective 
memory chips in the sample. Find $Var(X)$

a. if the sample is taken with replacement?

    In this setting $X$ is a binomial random variable with $n = 20$ and $\pi = 0.04$. By Theorem 7.3, 
    \begin{align*}
    Var(X) &= n \times \pi \times (1- \pi) \\
    &= 20 \times 0.04 \times 0.96.
    \end{align*}
    
    This is calculated below.
    
    ```{r}
    20*0.04*0.96
    ```

b. if the sample is taken without replacement? 

    Here the PMF of $X$ is given by
    $$f(x) = \frac{\binom{4}{x} \binom{96}{20 - x}}{\binom{100}{20}} \quad x = 0,1,2,3,4$$
    
    By Theorem 7.2, a shortcut formula for the variance is 
    
    \begin{align*}
    Var\left[X \right] &= E\left[X^2 \right] - \mu^2\\
    &= \sum\limits_{x} x^2 \cdot f(x) - \mu^2
    \end{align*}
    
    where $\mu = E\left[X \right]$. From last week, we know that $\mu = 0.8$. (Note: please look up solutions from section last Friday if you need a reminder)
    
    Therefore we need to find $E\left[X^2\right]$. This is done below. 
    
    ```{r}
    x <- 0:4
    fx <- choose(4,x)*choose(96,20-x)/choose(100,20)
    mu <- sum(x*fx); mu
    EX2 <- sum(x^2*fx ); EX2
    Var_X <- EX2 - mu^2; Var_X
    
    ```
    
    As we can see, when a sample is drawn without replacement, the variability is lower. This makes sense since there are fewer items to choose from on each subsequent draw which reduces the variability.
    

  


3. If $X$ denotes a temperature of a randomly selected day recorded in degrees Fahrenheit, then $Y = \frac{5}{9} X - \frac{160}{9}$ is the corresponding temperature in degrees Celsius. If the standard deviation for $X$ is 15.7$^{\circ} F$, what is the standard deviation of $Y$?

   By Lemma 7.4, we have the result that 
   $$Var(Y) = \left(\frac{5}{9} \right)^2 Var(X).$$
   Since $Var(X) = 15.7^2$ we can say that
   $$Var(Y) = \frac{25 \times 15.7 \times 15.7}{81}$$ and
   $$SD(Y) = \sqrt{Var(Y)} = \frac{5 \times 15.7}{9}.$$
   
    
4. (Chebychev) Suppose $X$ is a random variable with mean and variance both equal to 20. What can be said about $P( 0 < X < 40)$?

    Chebychev's inequality says that 
    $$P(|X - \mu| \geq k ) \leq \frac{\sigma^2}{k^2}.$$ 
    
    Here $\mu = 20$ and $\sigma^2 = 20$. Therefore by Chebychev's rule we can say
    \begin{align}
    P(|X - 20| \geq k) \leq \frac{20}{k^2} \label{one}
    \end{align}
   
 
    Now it is important to note that the event $0 < X < 40$ can be written as 
    \begin{align*}
    0 < X < 40 &= (0 - 20) < X - 20 < (40 - 20), \\
    &= -20 < X - 20 < 20, \\
    &= |X - 20| < 20
    \end{align*}
    
    Therefore 
    \begin{align*}
     P( 0 <  X < 40)  &= P(|X - 20| < 20) \\
     &= 1 - P(|X - 20| \geq 20) \tag{law of complements}\\
     &\geq  1 -\frac{20}{20^2} 
     \end{align*}
     
     where the last inequality follows from equation (\ref{one}) by reversing the direction of the inequality and with $k = 20$.
     
     Therefore we can say
     $$P(0 < X < 40) \geq \frac{19}{20}$$
     
     
    
  