---
title: "Problem Section 1 KEY"
date:  "Onboarding"
graphics: yes
output: 
  pdf_document:
        highlight: null
        pandoc_args: ["--syntax-highlighting=tango"]
header-includes:
    - \usepackage{amsmath, amssymb}
    - \usepackage{framed}\definecolor{shadecolor}{rgb}{0.949,0.949,0.949}
---


* * * 

\begin{shaded}

\textbf{Learning Outcomes}

The problems are designed to build conceptual understanding and problem-solving skills. The emphasis is on learning to find, evaluate and build confidence.
The specific tasks include: 

   - Describe events of interest using set notation.
     
   - Enumerate and count permutations and combinations
  
   - Back up and support work with relevant explanations


\end{shaded}


* * *
### Exercises 

1. For each of the following experiments, describe the sample space $S$.

a. An arborist counts the number of insect-damaged leaves on a plant with $N$ leaves. Characterize the possible outcomes.
    
   $S = \{0,1,2,  \dots, N\}$ 
	 
b. Quality control records the proportion of defectives in a shipment of $N$ electronic components. Characterize the possible outcomes.
	
	 $S = \{0,\frac{1}{N},\frac{2}{N},\dots,1\}$

2. Suppose a person is randomly selected from a large city. Let $A$ be the event that the randomly selected person is male, $B$ the event that the person is under 30 and $C$ that the person speaks French. Write the following verbal description of events using set operations on $A$, $B$, $C$. Venn diagrams might help you to figure out answers to this.  

a. A male at least 30 years old
    
    $A \cap B^c$
	
b. A female under 30 who speaks French
    
    $A^c \cap B \cap C$
	
c. A male who is either under 30 or speaks French

    $(A \cap B) \cup (A \cap C)$ or $A \cap (B \cup C)$


3. Use the multiplication rule of counting to find the number of

a. integers from 100 to 999 with distinct digits.

    We saw this problem in class, but now we need to count the number of three digit numbers with distinct digits. 
    
    First digit: 9 choices (1 - 9)
  
    Second digit: Still 9 choices (0 + remaining 8)
    
    Third digit: 8 choices.
    
    Therefore we have $9 \times 9 \times 8$ choices by the multiplication rule for counting.
    


b. odd integers from 100 to 999 with distinct digits.

    Here it is more efficient to first make the choice for the third digit, then the first and then the second. Think about why.
    
    Third digit: 5 choices (1,3,5,7,9)

    First digit: 8 choices (even numbers 2,4,6,8 and 4 odd numbers).

    Second digit: 8 choices (7 remaining numbers plus 0)
    
    Therefore we have $5\times 8 \times 8$ choices by the multiplication rule for counting.
    

4. A tourist wants to visit six of America's ten largest cities. 

a. In how many ways can she do that if the order of her visits matters to her. Calculate the answer in the code chunk provided below. 


    Here we want to know how many ways we can order a set of 6 cities from 10. By the multiplication rule, we have $10 \times 9 \times 8 \times 7 \times 6 \times 5$ or $\frac{10!}{4!}$
    
    ```{r ex4a}
    #number of itineraries if order matters
    factorial(10)/factorial(4)
    ```

    
b. In how many ways can she do that if the order of her visits does not matter to her. Create a small code chunk to calculate the answer. 

    Here we want to know how many ways we can choose 6 cities from 10. There are $\binom{10}{6}$ combinations which is calculated below.

    ```{r ex4b}
    #number of itineraries if order does not matter
    choose(10,6)
    ```
